---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# withDT

`withDT()` is called to use *data.table* syntax for one call.

## Installation

Install with:

``` r
# install.packages("devtools")
devtools::install_github("moodymudskipper/withDT")
```

## Assignments by reference

An important particularity of `withDT()` is that assignments are never done by
reference. Though limiting it avoids the confusion and unintended behaviors
that might come with them. The syntax of these assignments is still supported
but will return a copy. In order to fail explicitly whenever that syntax is used, 
the argument `lock` can be set to `TRUE`.

## Examples

```{r}
library(withDT)
```

```{r, echo = FALSE}
environment(withDT) <- .GlobalEnv
```


We can use standard *data.table* syntax and the output will have the same class
as the input.

```{r}
iris2 <- withDT(iris[, .(meanSW = mean(Sepal.Width)), by = Species][,cbind(.SD, a =3)])
iris2
class(iris2)
```

The following also works, but wouldn't have the same output with standard
*data.table* code due to the way assignments by reference are handled.

However we think in these cases the behavior of `withDT()` is more likely 
to be expected.

```{r}
iris3 <- withDT(iris[, .(meanSW = mean(Sepal.Width)), by = Species][,a:=3])
identical(iris2,iris3)
```


We see that `iris` wasn't modified

```{r}
class(iris)
names(iris)
```


To trigger an error when this syntax is used we can set `lock` to `TRUE`.

```{r}
try(iris4 <- withDT(lock=TRUE,iris[, .(meanSW = mean(Sepal.Width)), by = Species][,b:=3]))
```

## Other benefits

* Leverage compact *data.table* syntax, and speed to a degree
* Clearly isolate *data.table* code to avoid confusion due to hybrid syntax
* Be sure that no `data.table` object is created in calling environment, avoiding
  potential confusion
* Keep the class of `x` (regular data frame, tibble or other), with some
  exceptions detailed below
* *data.table* is not attached so there is no risk of masking functions from
  other packages such as *lubridate* (11 conflicting functions), 
  *dplyr* (4 conflicting functions) or *purrr* (1 conflicting function)

## Some caveats

* Copies are created for every call of `[`, which might be costly in some
  cases, if optimization is necessary we suggest using `data.table::setDT()`
  followed by idiomatic *data.table* syntax
* Attributes other than class are stripped
* Classes `grouped_df` and `rowwise_df`, created respectively
  by `dplyr::group_by()` and `dplyr::rowwise()` are stripped
* When using syntax of assignment by reference and `lock` is `FALSE` (default),
  some expressions won't be equivalent to the data.table code, see examples
* At the moment `withDT()` fails in markdown unless we assign it to the global
environment.

A workaround to the latter is to include the following chunk at the top of
your document : 

```{r, echo = FALSE, comment = NA}
cat("```{r, echo = FALSE}", "environment(withDT) <- .GlobalEnv","```", sep = "\n")
```
